Nested sampling with two objective functions

### 2015-11-04 Paper/ms.tex
tweaks

Nested Sampling \citep[NS][]{skilling} is an effective and popular
Monte Carlo algorithm which can solve a wide range of problems in Bayesian
computation and statistical mechanics
\citep{2009arXiv0906.3544P, 2014PhRvE..89b2302P, 2015arXiv150303404B}.
In a Bayesian inference problem with unknown parameters denoted collectively
by a vector $\xx$, the
posterior distribution for the parameters given data $\data$ (and prior
information $\II$) is:
\begin{eqnarray}
p(\xx | \data, \II) &=&
\frac{p(\xx | \II)p(\data | \xx, \II)}{p(\data | \II)}\\
&=& \frac{\pi(\xx)L(\xx)}{Z}
\end{eqnarray}
where $\pi(\xx)$ is the prior distribution, $L(\xx)$ is the likelihood
function, and $Z$ is the normalising constant, usually known as the
``marginal likelihood'' or the ``evidence'':
\begin{eqnarray}
Z &=& \int \pi(\xx) L(\xx) \, \dx.\label{eqn:evidence}
\end{eqnarray}
The main goal of NS is to approximate $Z$.
This is done by drawing particles from the
prior $\pi(\xx)$ and successively imposing constraints on the value of
the likelihood $L(\xx)$ that compress the prior mass by an
approximately-known factor. Like other related algorithms such as those
based on ``annealing'', NS
moves through a sequence of probability distributions beginning from the
prior. This sequence of distributions is defined by
\begin{eqnarray}
p(\xx; \ell) &=& \frac{1}{X(\ell)}\pi(\xx)\mathds{1}\left(L(\xx) > \ell\right).
\end{eqnarray}

In addition to the marginal likelihood $Z$, NS also allows us
to approximate properties of the posterior $\pi(\xx)L(\xx)/Z$.
NS is required because Equation~\ref{eqn:evidence}
is the expected value of $L$ with respect to a very heavy-tailed distribution
(the prior for the $L$ value implied by $\pi(\xx)$). Hence, there is a strong
connection between NS and ideas from rare event simulation \citep{walter}.
NS also allows us to calculate the properties of any other distribution that
is in some sense intermediate
between the prior and the posterior. For example, we might be interested in
a ``power posterior'' where the likelihood is raised to a power $\beta$:
\begin{eqnarray}
p(\xx; \beta) &=& \frac{\pi(\xx)L(\xx)^\beta}{Z(\beta)}\label{eqn:power_posterior}
\end{eqnarray}
The normalisation and posterior samples from this distribution can be obtained
from the original NS run by re-weighting the output according to $L(\xx)^\beta$
instead of the usual $L(\xx)$ used to obtain the posterior.
An example application, in the case of
a Bayesian inference problem with a ``gaussian noise'' likelihood assumption,
computing $p(\xx; \beta)$ for $\beta \neq 1$ allows us to explore what the
posterior distribution would have been if the noise variance had been greater.
This is different from including an extra parameter to allow the noise variance
to be greater, because NS allows you to test the consequences of values of the
variance that are very implausible given the data.

Alternatives to NS include methods based on ``annealing'', where a sequence of
distributions of the form of Equation~\ref{eqn:power_posterior} is used.
There are many different methods based on this idea, such as the popular
parallel tempering method \citep{pt}.
However, unlike annealing methods, NS continues
to work when the problem contains a phase transition.

@@

In some inference and
statistical mechanics problems, there are two or more scalar functions of
$\xx$ that are relevant. Suppose our prior is $\pi(\xx)$ as before, and
we obtain testable information that fixes the expected values of two scalar
functions of $\xx$, $L_1(\xx)$ and $L_2(\xx)$. It is well known that the
updated probability distribution that takes into account the constraints is
of the ``canonical'' form:
\begin{eqnarray}
p(\xx; \beta_1, \beta_2) &=& \frac{\pi(\xx)\exp\left[\beta_1L_1(\xx)+\beta_2L_2(\xx)\right]}
{Z(\beta_1, \beta_2)}
\end{eqnarray}
If we were only interested in a single canonical distribution, for example
with $\beta_1 = 0.3$ and $\beta_2 = 0.7$, we could estimate its normalising
constant and by running standard Nested Sampling with ``likelihood''
$L(\xx) = \exp\left[0.3L_1(\xx) + 0.7L_2(\xx)\right]$. However, usually we
are interested in a range of values for $\beta_1$ and $\beta_2$, and we
want to know the normalisation $Z(\beta_1, \beta_2)$, called the
partition function. This work describes
progress towards solving this class of problems while maintaining the benefits
of Nested Sampling, such as the ability to cope with first-order phase
transitions. The algorithm is designed for problems with two objective
functions, but should generalize to problems with more than two (though perhaps
not more than a few).

@@

@@

Imagine computing $Z(\beta_1, \beta_2)$ by looping over a grid of $\beta_2$
values and doing an NS run with respect to $L_1$ for each. For each
value of $\beta_1$ and $\beta_2$ that we're interested in, we can expect
a (marginal) ``one-sigma'' accuracy in
$\ln\left[Z(\beta_1, \beta_2)\right]$ of $\pm \sqrt{H(\beta_1, \beta_2)/N}$,
for a computational cost proportional to...

@@

The following argument is due to John Skilling. Consider Bayesian model selection
between two models $M_1$ and $M_2$. The Bayes Factor is the ratio of the
two marginal likelihoods:
\begin{eqnarray}
B = \frac{P(D | M_1)}{P(D | M_2)} = \frac{Z_1}{Z_2}
\end{eqnarray}
In large problems, the log of the Bayes Factor is usually more convenient:
\begin{eqnarray}
\ln B = \ln Z_1 - \ln Z_2.
\end{eqnarray}
Naively, we want $\ln B$ to an accuracy of about 0.1, so that we can reliably
quantify the strength of the evidence when the two models are similarly
supported by the data. But is this a realistic situation?
Prior to the data, we don't know the
value of $\ln B$. Rather, $\ln B$ has a prior distribution. We can find this
prior in terms of its CDF:
\begin{eqnarray}
P(\ln B \leq R) &=& \int P(D)P(\ln B \leq R | D) \, dD\\
&=& \int P(D) \mathds{1}\left(\ln B(D) \leq R\right) \, dD\\
&=& \int \sum_{i\in\{1,2\}}\left[P(M_i)P(D|M_i)\right] \mathds{1}\left(\ln B(D) \leq R\right) \, dD
\end{eqnarray}

>>>>

Nested Sampling \citep[NS][]{skilling} is an effective and popular
Monte Carlo algorithm which can solve a wide range of problems in Bayesian
computation and statistical mechanics
\citep{2009arXiv0906.3544P, 2014PhRvE..89b2302P, 2015arXiv150303404B}.
In a Bayesian inference problem with unknown parameters denoted collectively
by a vector $\xx$, the
posterior distribution for the parameters given data $\data$ (and prior
information $\II$) is:
\begin{eqnarray}
p(\xx | \data, \II) &=&
\frac{p(\xx | \II)p(\data | \xx, \II)}{p(\data | \II)}\\
&=& \frac{\pi(\xx)L(\xx)}{Z}
\end{eqnarray}
where $\pi(\xx)$ is the prior distribution, $L(\xx)$ is the likelihood
function, and $Z$ is the normalising constant, usually known as the
`marginal likelihood' or the `evidence':
\begin{eqnarray}
Z &=& \int \pi(\xx) L(\xx) \, \dx.\label{eqn:evidence}
\end{eqnarray}
The main goal of NS is to approximate $Z$.
This is done by drawing particles from the
prior $\pi(\xx)$ and successively imposing constraints on the value of
the likelihood $L(\xx)$ that compress the prior mass by an
approximately-known factor. Like other related algorithms such as those
based on `annealing', NS
moves through a sequence of probability distributions beginning from the
prior. This sequence of distributions is defined by
\begin{eqnarray}
p(\xx; \ell) &=& \frac{1}{X(\ell)}\pi(\xx)\mathds{1}\left(L(\xx) > \ell\right).
\end{eqnarray}

In addition to the marginal likelihood $Z$, NS also allows us
to approximate properties of the posterior $\pi(\xx)L(\xx)/Z$.
NS is required because Equation~\ref{eqn:evidence}
is the expected value of $L$ with respect to a very heavy-tailed distribution
(the prior for the $L$ value implied by $\pi(\xx)$). Hence, there is a strong
connection between NS and ideas from rare event simulation \citep{walter}.
NS also allows us to calculate the properties of any other distribution that
is in some sense intermediate
between the prior and the posterior. For example, we might be interested in
a `power posterior' where the likelihood is raised to a power $\beta$:
\begin{eqnarray}
p(\xx; \beta) &=& \frac{\pi(\xx)L(\xx)^\beta}{Z(\beta)}\label{eqn:power_posterior}
\end{eqnarray}
The normalisation and posterior samples from this distribution can be obtained
from the original NS run by re-weighting the output according to $L(\xx)^\beta$
instead of the usual $L(\xx)$ used to obtain the posterior.
An example application, in the case of
a Bayesian inference problem with a `gaussian noise' likelihood assumption,
computing $p(\xx; \beta)$ for $\beta \neq 1$ allows us to explore what the
posterior distribution would have been if the noise variance had been greater.
This is different from including an extra parameter to allow the noise variance
to be greater, because NS allows you to test the consequences of values of the
variance that are very implausible given the data.

Alternatives to NS include methods based on `annealing', where a sequence of
distributions of the form of Equation~\ref{eqn:power_posterior} is used.
There are many different methods based on this idea, such as the popular
parallel tempering method \citep{pt}.
However, unlike annealing methods, NS continues
to work when the problem contains a phase transition.

@@

In some inference and
statistical mechanics problems, there are two or more scalar functions of
$\xx$ that are relevant. Suppose our prior is $\pi(\xx)$ as before, and
we obtain testable information that fixes the expected values of two scalar
functions of $\xx$, $L_1(\xx)$ and $L_2(\xx)$. It is well known that the
updated probability distribution that takes into account the constraints is
of the `canonical' form:
\begin{eqnarray}
p(\xx; \beta_1, \beta_2) &=& \frac{\pi(\xx)\exp\left[\beta_1L_1(\xx)+\beta_2L_2(\xx)\right]}
{Z(\beta_1, \beta_2)}
\end{eqnarray}
where $\beta_1$ and $\beta_2$ are two `inverse temperatures'.
If we were only interested in a single canonical distribution, for example
with $\beta_1 = 0.3$ and $\beta_2 = 0.7$, we could estimate its normalising
constant and by running standard Nested Sampling with `likelihood'
$L(\xx) = \exp\left[0.3L_1(\xx) + 0.7L_2(\xx)\right]$. However, usually we
are interested in a range of values for $\beta_1$ and $\beta_2$, and we
want to know the normalisation $Z(\beta_1, \beta_2)$, called the
partition function. This work describes
progress towards solving this class of problems while maintaining the benefits
of Nested Sampling, such as the ability to cope with first-order phase
transitions. The algorithm is designed for problems with two objective
functions, but should generalize to problems with more than two (though perhaps
not more than a few).

@@

The true partition function (computed numerically) and the Kullback-Leibler
divergence from the prior to the canonical distribution are shown in
Figure~\ref{fig:truth} as a function of two temperatures.

@@

Imagine computing $Z(\beta_1, \beta_2)$ by looping over a grid of $\beta_2$
values and doing an NS run with respect to $L_1$ for each. For each
value of $\beta_1$ and $\beta_2$ that we're interested in, we can expect
a (marginal) `one-sigma' accuracy in
$\ln\left[Z(\beta_1, \beta_2)\right]$ of $\pm \sqrt{H(\beta_1, \beta_2)/N}$,
for a computational cost proportional to...

@@

The following argument is due to John Skilling. Consider Bayesian model selection
between two models $M_1$ and $M_2$. The Bayes Factor is the ratio of the
two marginal likelihoods:
\begin{eqnarray}
B = \frac{P(D | M_1)}{P(D | M_2)} = \frac{Z_1}{Z_2}
\end{eqnarray}
where $Z_1 = \int \pi(\xx_1)L(\xx_1) \, d\xx_1$ and
$Z_2 = \int \pi(\xx_2)L(\xx_2) \, d\xx_2$.
In large problems, the log of the Bayes Factor is usually more convenient:
\begin{eqnarray}
\ln B = \ln Z_1 - \ln Z_2.
\end{eqnarray}
Naively, we want $\ln B$ to an accuracy of about 0.1, so that we can reliably
quantify the strength of the evidence when the two models are similarly
supported by the data. But is this a realistic situation?
Prior to the data, we don't know the
value of $\ln B$. Rather, $\ln B$ has a prior distribution. We can find this
prior via its CDF:

{\bf Need thought here}
{\bf Thermodynamic integral might be helpful?}
